%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter6.tex
%% NOVA thesis document file
%%
%% Chapter with lots of dummy text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter6.tex}%

\chapter{OCAML-FLAT - Implementation}
\label{cha:implementation}

OCaml-FLAT is one of the two main components that are part of the existing structure of this project. It consists of the core logic of the tool and 
is implemented entirely in OCaml. In this chapter, we will address OCaml-FLAT, which contains the main operations implemented for attribute grammars.

OCaml-FLAT is a library composed of several modules, and each module is built as a class. This library, up to the moment of writing this document, 
implements pushdown automata, regular expressions, context-free grammars, as well as some supporting modules, such as operations on JSON files, errors, 
parsers, among others.

For the implementation of attribute grammars in OCaml-FLAT, the central goal is to couple a precise static validation phase with a deterministic evaluation 
strategy over parse trees. The library represents an attribute grammar as a set of rules with equations and conditions attached to each production, and it 
enforces a good construction before any evaluation can proceed. Validation checks range from context-free sanity (reusing CFG validation) to attribute 
typing, direction constraints (inherited vs.\ synthesized), and global acyclicity of attribute dependencies. Only after this pipeline succeeds do we 
execute the attribute equations over a parse tree.

Two aspects are key to the design. First, expression typing and equation validation ensure each assignment is meaningful and referentially correct. 
Second, the evaluation order guarantees that inherited attributes of a child are computed before descending into that child, and synthesized attributes 
of the head are computed after its children return---yielding a left-to-right, node-local schedule that is predictable and terminating in the absence of 
dependency cycles. Together, these choices produce a robust and explainable evaluator without relying on heuristic fixpoint closures.

Typing proceeds by a straightforward rule: attributes are associated to a base type inferred from their name---by default integers, with \texttt{"s"} 
denoting string and \texttt{"b"} denoting boolean---while constants and operators are type checked accordingly. Binary operators are admitted only when 
operand types align (e.g., \texttt{+} on \texttt{int} or \texttt{string}, \texttt{-}, \texttt{*}, \texttt{/} on \texttt{int}, comparisons across equal 
types), and any mismatch is flagged at validation time. Applications of attributes use \texttt{(attr, (var, i))}, where \texttt{i} identifies the head 
or a specific child occurrence; an index normalization step resolves the default case and unambiguous references.

On the left-hand side of equations, the direction discipline is enforced: assignments targeting the head must write synthesized attributes, whereas 
assignments targeting a child must write inherited attributes. Indices are checked against the rule's body to prevent out-of-range references, and 
all variables must belong to the grammar's variable set. If types on both sides of an equation do not match, validation fails. Conditions are validated 
separately and must type check to boolean.

At evaluation time, the library models a local environment as a list of nodes, headed by the current symbol and followed by its children. Each node 
carries the set of evaluated attributes for that symbol occurrence. Attribute applications resolve by locating the appropriate node---head or child 
occurrence---and reading the requested attribute value. Updates are performed by replacing the targeted attribute binding on the designated node, 
preserving immutability of the structure. A stable ordering of equations at the head and per child ensures determinism of side-effect-free updates 
within a node.

The evaluation strategy is left-to-right on children. For each child, all inherited equations targeting that child are applied in the head's environment 
before the recursive descent. The child is then evaluated. Once all children have been processed, equations that target the head (i.e., synthesized 
attributes) are evaluated, and node-local conditions are checked. This yields a two-phase per-node schedule---child inherited first, head synthesized 
later---that matches the intuition for attribute flow while avoiding ad hoc closures over \(\varepsilon\)-like effects.

Global correctness is guarded by a dependency graph construction. For every equation, the implementation identifies the destination attribute occurrence 
(head or specific child) and the set of attribute references appearing on the right-hand side. From these, it builds a directed graph whose nodes 
are ``(symbol, occurrence, attribute)'' triplets and whose edges represent ``RHS depends on LHS.'' A depth-first search detects cycles; if any cycle 
exists, the grammar is rejected as having circular attribute dependencies. This check is integrated into the overall \texttt{validate} routine, 
preventing non-terminating evaluations at runtime.

The attribute grammar layer interacts with the context-free layer. Conversions are provided in both directions: an attribute grammar can be projected to 
a CFG by erasing attributes, and a CFG can be lifted to an attribute grammar with empty attribute sets and equation blocks. Word acceptance may, when 
appropriate, delegate to the CFG acceptance procedure, while a separate routine attempts attribute evaluation on a given parse tree, reporting success 
or failure. A generator uses the CFG projection to enumerate terminal strings up to depth and count limits, which is useful for testing attribute 
specifications independently of a parser.

Reference semantics for attribute occurrences are explicit. Index \texttt{0} denotes the head; index \texttt{-1} is normalized to the head when the variable 
equals the head, and otherwise to the first occurrence of the named child; positive indices select the \(k\)-th matching child. Occurrence computations and 
equation ordering are derived from the rule body, ensuring that references are stable and reproducible even when the same child symbol appears multiple 
times. This removes ambiguity at both validation and evaluation time.

\section{Data types created for attribute grammars in Ocaml} 

The main data types defined to simplify the manipulation of attribute grammars are: \texttt{t}, which represents an attribute grammar (alphabet, 
variables, synthesized and inherited attributes, initial variable, and set of rules); \texttt{rule}, which models a production with \emph{head}, \
emph{body}, a set of \texttt{equations}, and a set of \texttt{conditions}; and \texttt{equation}, which represents an assignment between 
\texttt{expression}s. Additionally, \texttt{expression} and \texttt{value} describe the expressions and constants used in equations; 
\texttt{condition} represents boolean predicates; and \texttt{attribute}, \texttt{attributes}, and \texttt{attrArg} handle the identification of 
attributes and their occurrences.

\begin{verbatim}
type attribute = symbol
    type attributes = attribute set
    type attrArg = variable * int
    type value =
        | Int of int
        | String of string
        | Bool of bool
    type expression =
        | Const of value
        | Apply of attribute * attrArg (* l(A2) *)
        | Expr of string * expression * expression
    type equation = expression * expression
    type equations = equation set
    type condition = expression 
    type conditions = condition set

    type rule = {
        head : variable;
        body : word;
        equations : equations;
        conditions : conditions
    }
    type rules = rule set

    type t = {
        alphabet : symbols;
        variables : variables;
        synthesized : attributes;
        inherited : attributes;
        initial : variable;
        rules : rules
    }
\end{verbatim}

\section{Operations implemented} 

This section presents the design, static checks, and evaluation semantics of the attribute-grammar component. 
The implementation supports typed attributes (integers, strings, booleans), distinguishes inherited and synthesized attributes, validates equations and 
conditions, constructs a dependency graph to detect cycles, evaluates attributes over parse trees with a deterministic order, and supplies conversions to 
and from context-free grammars (CFG).


\subsection{The \texttt{calcAttributes} fuction} 

\texttt{calcAttributes} evaluates all attributes in a parse tree according to a given attribute grammar. 
It ensures that each child node receives its inherited attributes before that child is evaluated, and that the parent node computes its synthesized attributes 
only after all children have been evaluated. 
The function returns a tree with the same shape, but with attribute maps updated on the head node and on every descendant. 


\begin{verbatim}
let rec calcAttributes (ag : t) (pt : parseTree) : parseTree =
    match pt with
    | Leaf _ ->
        pt

    | Node (((head_sym, _) as head), children) ->
        let rule = getRootRule ag (Node (head, children)) in
        let children' = eval_children_left_to_right ag head rule.equations children in
        let head_eqs = set_filter (eq_targets_head head_sym) rule.equations in
        let env2 = head :: List.map getRoot children' in
        let env3 = apply_equations_at_head head_sym env2 head_eqs in
        check_conditions_at_node head_sym rule.conditions env3;

        match env3 with
        | (new_head_sym, new_head_evs) :: child_envs ->
            let children'' = zip_update_children children' child_envs in
            Node ((new_head_sym, new_head_evs), children'')
        | [] ->
            failwith "calcAttributes: empty environment after synthesized equations"
\end{verbatim}


\subsubsection{Inputs and Output}
The function takes an attribute grammar \texttt{ag} and a parse tree \texttt{pt}. If \texttt{pt} is a leaf, it is returned unchanged. 
If \texttt{pt} is an internal node, the function computes attributes for the node and for all its children, returning a structurally identical tree
whose nodes carry the evaluated attribute bindings. 

\subsubsection{Evaluation Procedure at an Internal Node}
Let the current node be \texttt{Node((head\_sym, head\_evs), children)}.
\begin{enumerate}
  \item \textbf{Rule identification.} The function determines the unique grammar rule whose head symbol equals \texttt{head\_sym} and whose body matches the 
sequence of child symbols at this node. If no rule matches, a diagnostic is raised that includes the encountered head/body and the candidate rules 
that share the same head. This step fixes the set of equations and conditions that apply at the node. 

  \item \textbf{Inherited attributes for children (left-to-right).} Children are processed strictly from left to right. For the current child occurrence, 
the function builds an \emph{environment} that consists of the head node, the already processed children (in order, with any attributes computed so far), 
and the remaining children (not yet processed). From the node's rule, it selects only those equations that assign \emph{inherited} attributes to this specific 
child occurrence (identified by symbol and 1-based occurrence index), and applies them to the environment in a stable order. The child's root is updated with these inherited 
values, and then \texttt{calcAttributes} is called recursively on that child to evaluate its subtree. This guarantees that a child's inherited attributes may depend on the parent 
and on earlier siblings, but not on later siblings. 

  \item \textbf{Synthesized attributes for the head.} Once all children are fully evaluated, the function gathers only the equations that target the head 
(i.e., assign \emph{synthesized} attributes). It constructs a head-first environment containing the head node followed by the now-evaluated child roots and 
applies those equations in a deterministic, stable order, updating the head's attribute bindings. 

  \item \textbf{Condition checking.} The boolean conditions attached to the rule are evaluated in the same head-first environment. Each condition must evaluate 
to \texttt{true}. If a condition fails or evaluates to a non-boolean value, evaluation aborts with an error message that identifies the node at which it failed. 

  \item \textbf{Reconstruction.} Finally, the function rebuilds and returns the node by replacing the head and each child root with their updated attribute 
environments. The tree's topology is unchanged; only attribute maps are updated. 
\end{enumerate}

\subsubsubsection{Deterministic Order and Rationale}
The strict left-to-right order establishes a predictable flow of information: inherited attributes for child \( i \) may depend on the head and on children
\( 1..i-1 \), but never on \( i+1..n \). This sequencing avoids evaluation-time circularities and ensures each child observes a stable context when it is 
evaluated. A separate dependency-graph check elsewhere detects global cycles among equations. 

\subsubsection{Error Handling}
The surrounding code validates that attribute references target valid head/child positions, checks expression types (e.g., integer vs.\ string arithmetic and 
comparisons), and detects inconsistent environments (e.g., out-of-range child occurrence, head-symbol mismatches). Failures are reported with precise diagnostics 
at the point of detection; rule-selection mismatches list candidate rules with the same head, and condition failures indicate the node where the evaluation failed. 


\section{Local Auxiliary Logic Used by \texttt{calcAttributes}}
This section focuses on the local worker that drives left-to-right child evaluation for \texttt{calcAttributes}. In the implementation, \texttt{calcAttributes} 
delegates the child pass to \texttt{eval\_children\_left\_to\_right}, which defines a local recursive function \texttt{loop}. 
This \texttt{loop} is the auxiliary routine that realizes the stepwise propagation of inherited attributes to each child and then triggers 
recursive evaluation of that child. 

The call to \texttt{eval\_children\_left\_to\_right} is the point where the local auxiliary function \texttt{loop} is used. 

The \texttt{loop} function above is the auxiliary worker for \texttt{calcAttributes}. 

\subsection{Role of the Auxiliary \texttt{loop} Function}
\textbf{Purpose.} \texttt{loop} performs a left-to-right sweep over the node's children. For each child in order, it computes and writes that child's inherited 
attributes, recursively evaluates the child subtree, and carries the result forward before proceeding to the next child. 
This realizes the required dependency discipline: a child may depend on the head and on previously evaluated siblings, but not on later siblings. 

\textbf{Interface (implicit via parameters).} The function carries three accumulators:
\begin{itemize}
  \item \texttt{processed}: the list of children already evaluated; kept in reverse order for efficient cons operations. These are fully evaluated subtrees. 
  \item \texttt{remaining}: the list of children yet to process, with the next child at the head. 
  \item \texttt{remaining\_o}: the list of $(\textit{symbol}, \textit{occurrence})$ pairs corresponding one-to-one with \texttt{remaining}, used 
  to target inherited-equation assignments to the correct child occurrence. 
\end{itemize}

\textbf{Key invariants.}
\begin{enumerate}
  \item \texttt{processed @ remaining} is a permutation of the original \texttt{children}, preserving relative order between the still-unseen suffix and the 
  already-evaluated prefix. The invariant is maintained by consing the newly evaluated child onto \texttt{processed} and dropping it from \texttt{remaining}. 
  \item \texttt{remaining\_o} matches \texttt{remaining} position-by-position; each pair $(s,k)$ gives the symbol and $k$-th occurrence index of the corresponding 
  child among all children of this node. This allows precise selection of inherited equations and of the correct child node in the environment. 
  \item Every tree in \texttt{processed} is fully evaluated (its inherited attributes were set before the recursive call, and its synthesized attributes 
  and conditions were handled inside that recursive \texttt{calcAttributes}). 
\end{enumerate}

\subsection{Step-by-Step Operation of \texttt{loop}}
For the case \verb|child :: rest, occ :: occs_rest|:
\begin{enumerate}
  \item Compute \texttt{processed\_nodes} as the roots of already-evaluated children (in original order). Then build the evaluation environment for this step:
  \[
    \texttt{env\_before} =
    \texttt{head} \;::\;
    \texttt{processed\_nodes} \; @ \;
    \texttt{(roots of (child :: rest))}.
  \]
  This environment exposes the head, earlier siblings, the current child, and later siblings to the equations to be applied. 

  \item Select only inherited-attribute equations that target the current occurrence \texttt{occ}:
  \[
    \texttt{inh\_eqs} =
    \{\, e \in \texttt{eqs} \mid \texttt{eq\_targets\_child head\_sym occ}(e)\,\}.
  \]
  Apply them in a stable order at the head of \texttt{env\_before}:
  \[
    \texttt{env\_after\_inh} =
    \texttt{apply\_equations\_at\_head head\_sym env\_before inh\_eqs}.
  \]
  The effect is to update exactly the target child’s inherited attributes in the environment. 

  \item Extract the updated root node for the current child occurrence from the environment tail using \texttt{pick\_child\_node\_by\_occ occ}, and write 
  it back into the child tree with \texttt{updateRoot}:
  \[
    \texttt{child\_with\_inh} = \texttt{updateRoot child child\_node\_after\_inh}.
  \]
  If the environment is malformed or the occurrence is not found, a descriptive failure is raised. 

  \item Recursively evaluate the child subtree:
  \[
    \texttt{child\_evaluated} = \texttt{calcAttributes ag child\_with\_inh}.
  \]
  This computes the child's own synthesized attributes and checks its local conditions. 

  \item Advance the loop:
  \[
    \texttt{loop (child\_evaluated :: processed) rest occs\_rest}.
  \]
  The measure \(|\texttt{remaining}|\) strictly decreases, ensuring termination. 
\end{enumerate}

When \texttt{remaining} and \texttt{remaining\_o} are both empty, the loop returns \texttt{List.rev processed}, i.e., the fully evaluated children in original 
left-to-right order. Any other shape indicates a mismatch between children and occurrence annotations and triggers an explicit failure. 

\subsection{Why This Auxiliary is Necessary}
The local \texttt{loop} encodes the evaluation discipline that inherited attributes of child \(i\) may depend on the head and on the already evaluated 
siblings \(1..i-1\), but not on later siblings. It does so by constructing \texttt{env\_before} at each step, selecting inherited equations for exactly 
one occurrence, and committing the results before recursing to the next child. This realizes a deterministic, left-to-right strategy and prevents 
accidental dependence on future siblings during inherited-attribute computation. 

\subsection{Relation Back to \texttt{calcAttributes}}
After \texttt{loop} has produced the fully evaluated children, \texttt{calcAttributes} proceeds to the parent's synthesized pass and condition checking. 
The correctness of those later phases relies on the contract guaranteed by \texttt{loop}: all children are available with their inherited attributes applied 
and their subtrees fully evaluated.


\subsection{Selected Functions: Explanations and Code}

\subsubsection{\texttt{getRootRule}}
Select the unique grammar rule that matches the current parse-tree node: the rule whose head symbol equals the node's root symbol and whose body 
(sequence of symbols) equals the sequence of the node's immediate children. If the node is a leaf, there is no applicable rule and the function fails. 
If no rule matches, the error message lists candidate rules that share the same head symbol to aid diagnosis. 

Input: the attribute grammar \texttt{ag} and a parse tree \texttt{pt}.  
Output: the matching rule \texttt{rule}.  
Failure cases: leaf nodes, or internal nodes whose children do not match any rule body for the node's head symbol. 

For an internal node, the head symbol is taken from the node's root and the body is the list of child-root symbols. The function searches \texttt{ag.rules} for 
a rule whose head and body match. If none matches, it constructs a readable error with candidate alternatives. 

\begin{verbatim}
let getRootRule (ag: t) (pt: parseTree): rule =
  match pt with
  | Leaf _ ->
      failwith "getRootRule: leaf has no rule"
  | Node (_, children) ->
      let head = getRootSymbol pt in
      let body = List.map getRootSymbol children in
      try Set.find (fun r -> r.head = head && r.body = body) ag.rules with
      | _ ->
          let candidates =
            ag.rules
            |> SetUtil.to_list
            |> List.filter (fun r -> r.head = head)
            |> List.map (fun r ->
                 Printf.sprintf "- %s -> %s"
                   (symb2str r.head)
                   (String.concat " " (List.map symb2str r.body)))
            |> String.concat "\n"
          in
          failwith (Printf.sprintf
            "getRootRule: no rule matches head=%s, body=[%s]\nCandidates with same head:\n%s"
            (symb2str head)
            (String.concat " " (List.map symb2str body))
            (if candidates = "" then "(none)" else candidates))
\end{verbatim}

\subsubsection{\texttt{apply\_equations\_at\_head}}

Apply a given set of equations to an evaluation environment whose first node must be the head node with a specific symbol. 
Equations are applied in a stable, deterministic order to avoid order-dependent results. The output is a new environment reflecting 
the updates introduced by these equations (e.g., inherited assignments to a specific child, or synthesized assignments to the head). 


Inputs: head symbol \texttt{head\_sym}, a non-empty environment \texttt{nodes} whose first node has symbol \texttt{head\_sym}, and an equation set \texttt{equations}.  
Output: an updated environment \texttt{node list}.  
Failure cases: empty environment or a head-symbol mismatch at the first node. 

The function first verifies that the environment is not empty and that its first node matches \texttt{head\_sym}. It then orders the equations using a 
stable ordering and folds over them, applying each equation with \texttt{eval} to produce the next environment. The fold result is the updated environment. 

\begin{verbatim}
let apply_equations_at_head
    (head_sym : symbol)
    (nodes : node list)
    (equations : equation Set.t)
    : node list =
  let () =
    match nodes with
    | [] ->
        failwith "apply_equations_at_head: empty environment (nodes)"
    | (hs, _) :: _ when hs = head_sym -> ()
    | (hs, _) :: _ ->
        failwith
          (Printf.sprintf
             "apply_equations_at_head: head symbol mismatch (expected %s, got %s)"
             (symb2str head_sym) (symb2str hs))
  in
  let eqs = stable_eq_order head_sym equations in
  List.fold_left
    (fun acc_nodes eq -> eval head_sym eq acc_nodes)
    nodes
    eqs
\end{verbatim}


\subsubsection{\texttt{check\_conditions\_at\_node}}

Evaluate the rule's boolean conditions at the current node's environment and ensure they all hold. A failing condition aborts evaluation with a message 
identifying the node's head symbol. 

Inputs: head symbol \texttt{head\_sym}, the set of conditions \texttt{conds}, and the current environment \texttt{env}.  
Output: \texttt{unit}.  
Failure cases: any condition evaluates to a non-boolean value or to \texttt{false}. 

Each condition is evaluated with \texttt{evaluate}; the result is coerced with \texttt{as\_bool}. If any evaluation fails to yield \texttt{true}, 
a failure is raised with a diagnostic mentioning the head symbol. 

\begin{verbatim}
let check_conditions_at_node
    (head_sym : symbol)
    (conds : condition Set.t)
    (env : node list)
    : unit =
  Set.iter (fun cond ->
    let v = evaluate head_sym cond env in
    if not (as_bool v) then
      failwith (Printf.sprintf
        "Condition failed at node %s"
        (symb2str head_sym))
  ) conds
\end{verbatim}


\subsection{The \texttt{accept}}

The \texttt{accept} receives
a word and an attribute grammar, and returns a boolean indicating whether the word
is accepted by the grammar. The function first converts the attribute grammar into an
equivalent context--free grammar and then relies on the acceptance mechanism of that
grammar.

\begin{verbatim}
let accept (ag: t) (w: word): bool =
    let cfg = ga_to_cfg ag in
    ContextFreeGrammarBasic.accept cfg w
\end{verbatim}


\subsection{Core Entities}

An attribute grammar instance $AG$ consists of terminal alphabet $\Sigma$, variable set $V$, an initial symbol $S \in V$, and a set of rules $R$. 
Each rule $r \in R$ has a head $A \in V$, a body $X_1\ldots X_k$ with $X_i \in \Sigma \cup V$, a finite set of equations, and a finite set of boolean 
conditions. Attributes are partitioned into inherited and synthesized; every symbol may carry a (finite) mapping from attribute names to values. Values 
are typed: \texttt{int}, \texttt{string}, and \texttt{bool}. A simple mapping assigns types to attribute names (e.g., attribute ``\texttt{s}'' is \texttt{string},
``\texttt{b}'' is \texttt{bool}, and any other is \texttt{int}). This per-attribute typing is used both in static validation and at run time during evaluation.

References to attributes in equations use the form \(\mathsf{attr}(X[i])\), where \(i=0\) denotes the rule head, \(i>0\) denotes the $i$-th occurrence of a 
variable in the body, and \(i=-1\) denotes the first occurrence when the variable appears at least once. An auxiliary normalization maps the default 
index \(-1\) to \(0\) when the referenced variable is the head symbol. These conventions allow uniform access to attributes of the head and of specific body occurrences.

\subsection{Static Validation of Equations and Conditions}

A valid equation is of the form \(\mathsf{attr}(X[i]) := e\). The validator confirms that the LHS applies to a legitimate symbol position: if \(i=0\), then \(X\) must be the rule head; if \(i>0\), then \(X\) must appear at least \(i\) times in the body; and if \(i=-1\), the referenced symbol must occur as required by the defaulting rule. Directionality is enforced: LHS targeting the head must name a synthesized attribute; LHS targeting a child must name an inherited attribute. Any mismatch (e.g., assigning an inherited attribute on the head) is rejected.

The static checker computes expression types compositionally. Constants carry their declared type. Attribute applications \(\mathsf{attr}(X[i])\) inherit the type associated with \(\mathsf{attr}\). Binary operators are type–checked against a fixed table: arithmetic (\texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}) require \texttt{int} operands (with \texttt{+} also overloaded for \texttt{string} concatenation); equality/inequality (\texttt{=}, \texttt{$<>$}) require operands of the same type; relational comparisons (\texttt{<}, \texttt{$<=$}, \texttt{>}, \texttt{$>=$}) apply to \texttt{int} and \texttt{string}. The parentheses operator is treated as projection. Inconsistent uses (e.g., comparing \texttt{bool} with \texttt{int}, or dividing by a non-\texttt{int}) are rejected. Conditions must type-check to \texttt{bool}.

The checker verifies that each referenced variable is declared, each attribute name belongs to either the inherited or synthesized sets, and child indices are in range with respect to the rule body. The default index \(-1\) resolves to the first occurrence when referencing children and to \(0\) when referencing the head symbol.

\subsection{Evaluation Model over Parse Trees}

Evaluation is defined on a parse tree that conforms to the underlying CFG structure (same head and body symbols per rule instance). Each tree node holds a symbol and an environment of evaluated attributes (a finite map from attribute names to values). Equations are applied in a deterministic order that guarantees that inherited attributes required by a child are computed before descending into that child, and that synthesized attributes of the head are computed after all children have been processed. Conditions are checked at the node once the required attributes are available. Failures abort evaluation at that node.

For each node, an environment list is constructed consisting of the head's evaluation map followed by the children's evaluation maps in left-to-right order. Attribute references are resolved by first normalizing the occurrence index, then selecting the correct evaluation map (head vs.\ child occurrence), and finally retrieving the attribute value from that map; missing attributes trigger a runtime error.

Inherited attributes for each child are computed left-to-right. For the $j$-th child, the evaluator forms an environment composed of the head, the already processed children (with their updated environments), and the remaining unprocessed children; it then filters the rule's equations to those targeting the specific child occurrence and applies them at the head context to produce the child's inherited values. Immediately after, evaluation recurses into that child. This procedure repeats for all children, preserving a deterministic left-to-right traversal.

After all children are evaluated, the evaluator selects the subset of equations that target the head and applies them in a stable, deterministic order. The ordering key first distinguishes head targets from child targets, then orders child targets by symbol name and occurrence index, and finally by attribute name; for head equations, the key ensures a predictable sequence by attribute name. The ordered fold over equations updates the head environment accordingly. Conditions attached to the rule are then evaluated in this final environment and must hold.

\subsection{Dependency Analysis and Cycle Detection}

The implementation constructs a dependency graph over attribute occurrences to reject cyclic specifications. Nodes of this graph are triples \((\mathsf{sym}, j, \mathsf{attr})\) denoting the \(j\)-th occurrence of symbol \(\mathsf{sym}\) within a rule instance (with \(j=0\) for the head), paired with an attribute name. For an equation \(\mathsf{attr}_L(X[i]) := e\), the construction adds edges from every attribute occurrence referenced in \(e\) to the target occurrence \((X, i, \mathsf{attr}_L)\). A standard depth-first search with recursion stack detection decides the presence of cycles. Any cycle yields a validation error that rejects the attribute grammar.

\subsection{Conversions to and from CFG}

Two structural conversions are provided. First, a forgetful map \(AG \rightarrow \mathrm{CFG}\) drops attributes and equations, preserving only the underlying production rules \((\text{head}, \text{body})\). Second, a CFG can be lifted to a empty attribute grammar by reusing its terminals, variables, initial symbol, and rules, while initializing the sets of inherited and synthesized attributes, equations, and conditions to be empty. These conversions support reuse of CFG infrastructure (e.g., acceptance testing) and serve as a baseline for generation procedures.

\subsection{Acceptance and Generation}

Grammar acceptance is delegated to the underlying CFG acceptance procedure obtained from the forgetful conversion. Hence, acceptance of a word depends on the context-free structure and not on attribute constraints. A separate predicate evaluates attributes over a given parse tree and returns success if attribute evaluation completes without errors and all conditions hold at every node.

A bounded breadth-first enumeration procedure produces terminal words derivable from the start symbol. The algorithm maintains a queue of forms together with their derivation depth. For each dequeued form, if it contains only terminals (after removing \(\epsilon\)), it is recorded as an output; otherwise, and provided a user-specified depth bound is not exceeded, the leftmost variable is expanded by all of its rule bodies, enqueuing the resulting sentential forms. A deduplication table avoids revisiting sentential forms already seen. The procedure returns up to a user-specified maximum number of words.

\subsection{Parse-Tree Interface and Rule Matching}

Evaluation over a parse tree requires that each internal node corresponds to a unique rule whose head equals the node's symbol and whose body equals the sequence of the children’s root symbols (after normalizing away \(\epsilon\)). If no rule matches, a diagnostic lists candidate rules sharing the same head to support debugging. Leaves carry empty attribute maps; internal nodes start with empty maps that are populated by evaluation. The tree structure is manipulated through utilities that extract the root node, its symbol, its children, and that update a child or the root with a new evaluation map.

\subsection{Condition Checking}

Rule-level conditions are boolean expressions evaluated at the node after inherited and synthesized attributes required by those expressions are available. Every condition must evaluate to \texttt{bool}; non-boolean results or failed conditions produce errors that stop evaluation. The implementation evaluates each condition to a value, coerces it to a boolean, and raises a failure if the result is false.

\subsection{Stable Equation Ordering and Determinism}

To ensure reproducibility and avoid accidental dependence on data-structure iteration order, equations are applied in a stable order determined by a total key. Targets to the head precede targets to children; among child targets, the order is by the child’s symbol name, then by the occurrence index (1-based), and then by attribute name. The resulting list is folded from left to right, updating the current environment at each step. This ordering interacts with the two-phase strategy (children first for inherited attributes, then head for synthesized) to yield deterministic outcomes.

\subsection{Auxiliary Analyses and Utilities}

Several utilities support the main algorithms. Rule bodies are normalized by removing \(\epsilon\); helper predicates recognize productions of the form \(A \rightarrow a\) and \(A \rightarrow BC\) for quick classification; a hash table indexes bodies by head symbol for efficient expansion during generation; and functions compute child- occurrence metadata used by the left-to-right inherited-attribute pass. Error messages are constructed with concrete values and symbols to aid debugging, including formatted renderings of values and descriptions of mismatched expectations.

\subsection{Complexity and Termination Considerations}

Static validation is linear in the size of the grammar and its equations plus the cost of dependency-cycle detection. Cycle detection runs in linear time in the size of the dependency graph (sum of nodes and edges). Attribute evaluation over a fixed parse tree is linear in the size of the tree times the number of equations per node, given that each equation and condition is evaluated once in a fixed order. The bounded generation procedure is exponential in the worst case with respect to depth, as typical for breadth-first exploration of derivations, but it is explicitly limited by user-set bounds on depth and number of words.
