%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter1.tex
%% NOVA thesis document file
%%
%% Chapter with introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter1.tex}%

\chapter{Introduction}
\label{cha:introduction}


\section{Context}
\label{sec:context}

The study of Formal Languages and Automata Theory (FLAT) is integral to 
Computer Engineering and Computer Science programs. However, due to its 
abstract and mathematical nature, students often struggle to fully grasp 
the concepts and models involved. Therefore, a tool that complements the 
study of FLAT theory by being practical, user-friendly, and visually 
oriented is highly beneficial.

In recent years, the OCamlFLAT/OFLAT tool has been developed at FCT-UNL, 
offering extensive support in this field. Despite its advanced state, 
with support for several key models, there are still many features that 
could enhance its functionality.

This dissertation aims to introduce new capabilities to OCamlFLAT/OFLAT, 
enhancing both the OCamlFLAT library and the OFLAT graphical interface.


\section{Objectives}
\label{sec:objectives}

The primary objective of this dissertation is to incorporate support for 
attribute grammars. Attribute grammars extend context-free grammars by 
allowing additional information (attributes) to be associated with 
production rules. These attributes are crucial for describing the 
semantic properties of the languages generated by the grammars.

To achieve this, the OCamlFLAT library needs to be extended to support 
attribute grammars. Additionally, graphical enhancements are required to 
enable the creation and editing of these grammars within the OFLAT 
platform.


\section{Document Structure}
\label{sec:document_structure}

This thesis is structured into multiple chapters, each covering essential aspects of the research, development, and implementation of the OCamlFLAT library and the OFLAT platform. The document follows a logical progression from theoretical background to practical applications. Below is an overview of its structure:

\begin{itemize}
    \item \textbf{Introduction}: Provides an overview of the dissertation's context, objectives, and structure. It sets the stage by highlighting the relevance of Formal Languages and Automata Theory (FLAT) in computer science and engineering, and outlines the goals of enhancing the OCamlFLAT/OFLAT platform.
    
    \item \textbf{Theory of Computation}: Lays the theoretical foundation for the study of FLAT models. It begins with an exploration of the Chomsky hierarchy and progresses into detailed discussions of the four language types (Type 3, 2, 1, and 0). The chapter also delves into model composition and decision problems, providing the theoretical background for the practical applications.
    
    \item \textbf{Grammar Tools}: Examines existing pedagogical tools that support the teaching and learning of FLAT theory, with a focus on three prominent systems: JFLAT, ANTLR, and CGM. Each tool is evaluated for its strengths and limitations in helping users understand and apply attribute grammars.
    
    \item \textbf{OCamlFLAT and OFLAT Systems}: Introduces the OCamlFLAT library and the OFLAT platform, discussing their respective roles in the toolset. The OCamlFLAT library forms the core logic of the system, implementing key FLAT models, while the OFLAT platform serves as the graphical user interface, enabling users to interact with these models through a web application.
    
    \item \textbf{Technology}: Covers the technical aspects of developing the OCamlFLAT/OFLAT systems, it have discussions on declarative programming in OCaml, functional programming paradigms, and OCaml/JavaScript interoperability, with an emphasis on the tools and methodologies employed to implement the platform's graphical interface.
    
    \item \textbf{Planning}: Outlines the detailed plan for the tasks to be accomplished throughout the project, including a timeline, key milestones, and a schedule for the work to be completed in the upcoming months.
\end{itemize}



